---
title: "Stable, GEV and GAt ARMA-APARCH models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stable, GEV and GAt ARMA-APARCH models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(GEVStableGarch)
```

GEVStableGarch fits, simulates and forecasts ARMA-GARCH/APARCH models whose
innovations follow a stable, GEV or GAt distribution, or any distribution you
define. This vignette walks through the workflow with stable innovations and
then builds a user defined distribution.

## The model

In the intercept form of Wuertz et al. (2006),

$$x_t = \mu + \sum_{i=1}^m a_i x_{t-i} + \sum_{j=1}^n b_j e_{t-j} + e_t, \qquad e_t = \sigma_t z_t,$$
$$\sigma_t^\delta = \omega + \sum_{i=1}^p \alpha_i (|e_{t-i}| - \gamma_i e_{t-i})^\delta + \sum_{j=1}^q \beta_j \sigma_{t-j}^\delta,$$

with $z_t$ i.i.d. from the innovation distribution, standardized to location
0 and scale 1. Stable innovations have infinite variance, so $\sigma_t$ is a
conditional *scale*, and $E|z|^\delta$ is finite only for $\delta < \alpha$.
In GARCH mode (`aparch = FALSE`) the power is fixed at 1 for stable
innovations and 2 for GEV and GAt.

## Stable AR(1)-APARCH(1,1)

A specification says which model to fit; a model adds parameter values.

```{r}
spec <- gs_spec(arma = c(1, 0), garch = c(1, 1), aparch = TRUE, dist = gs_stable())
spec
model <- gs_model(spec, mu = 0.0005, ar = 0.05, omega = 0.0002, alpha = 0.05,
                  gamma = 0.3, beta = 0.85, delta = 1.2,
                  dist_par = c(stable_alpha = 1.75, stable_beta = 0))
gs_stationarity(model)
```

The persistence $\beta + \alpha\, E(|z| - \gamma z)^\delta$ is below one, so
the model has a stationary scale level. Simulate and fit:

```{r}
x <- gs_sim(model, n = 1500, seed = 1)$y
fit <- gs_fit(x, spec)
fit
```

Standardized residuals and their probability integral transform under the
fitted innovation distribution (uniform when the model is right, and the
input of copula models):

```{r}
u <- residuals(fit, type = "pit")
ks.test(u, "punif")$p.value
```

## Forecasting and value at risk

One step ahead the forecast is exact: location, scale and quantiles of a
shifted and scaled stable law. The value at risk at level $p$ is minus the $p$
quantile.

```{r}
predict(fit, n_ahead = 1, level = c(0.01, 0.05))
```

Beyond one step the scale follows the APARCH recursion and the quantiles
come from simulated paths.

```{r}
predict(fit, n_ahead = 5, level = 0.01, n_sim = 5000, seed = 2)
```

In rolling windows, start each fit from the previous estimates and skip the
Hessian when only forecasts are needed:

```{r, eval = FALSE}
window <- 1000
var_1 <- numeric(0); prev <- NULL
for (t in window:(length(x) - 1)) {
  fit_t <- gs_fit(x[(t - window + 1):t], spec, start = prev, hessian = FALSE)
  prev <- coef(fit_t)
  var_1 <- c(var_1, -predict(fit_t, level = 0.01)$q_0.01)
}
```

## A user defined innovation distribution

Any distribution can be plugged in with `gs_dist()`. Every field that the
machinery needs is required, including when moments exist (`max_power`) and
the mean (`mean`, which must return `NA` explicitly if it does not exist).
The constructor checks the pieces against each other and stops with an error
naming the faulty field, so a mistake cannot silently distort a fit. Here the
normal distribution:

```{r}
norm <- gs_dist(
  name = "norm",
  log_density = function(z, par) dnorm(z, log = TRUE),
  random = function(n, par) rnorm(n),
  cdf = function(q, par) pnorm(q),
  quantile = function(p, par) qnorm(p),
  par_names = character(0), lower = numeric(0), upper = numeric(0), start = numeric(0),
  default_delta = 2,
  max_power = function(par) Inf,
  mean = function(par) 0)
```

A wrong declaration is caught immediately:

```{r, error = TRUE}
gs_dist(
  name = "norm",
  log_density = function(z, par) dnorm(z, log = TRUE),
  random = function(n, par) rnorm(n),
  cdf = function(q, par) pnorm(q),
  quantile = function(p, par) qnorm(p),
  par_names = character(0), lower = numeric(0), upper = numeric(0), start = numeric(0),
  default_delta = 2,
  max_power = function(par) Inf,
  mean = function(par) 0.5)
```

The new distribution gets the same fitting, stationarity and forecasting:

```{r}
spec_n <- gs_spec(garch = c(1, 1), dist = norm)
m_n <- gs_model(spec_n, mu = 0, omega = 0.1, alpha = 0.1, beta = 0.8)
fit_n <- gs_fit(gs_sim(m_n, n = 1000, seed = 3)$y, spec_n)
coef(fit_n)
predict(fit_n, n_ahead = 3, n_sim = 2000, seed = 4)
```

## References

Mittnik, S., Paolella, M. S. and Rachev, S. T. (2002). Stationarity of stable
power-GARCH processes. *Journal of Econometrics*, 106, 97-107.

Paolella, M. S. (1997). Tail estimation and conditional modeling of
heteroskedastic time series. PhD thesis, University of Kiel.

Wuertz, D., Chalabi, Y. and Luksan, L. (2006). Parameter estimation of ARMA
models with GARCH/APARCH errors: an R and SPlus software implementation.
*Journal of Statistical Software*.
