---
title: "Stable, GEV and GAt ARMA-APARCH models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stable, GEV and GAt ARMA-APARCH models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(GEVStableGarch)
```

GEVStableGarch fits, simulates and forecasts ARMA-GARCH/APARCH models whose
innovations follow a stable, GEV or GAt distribution, or any distribution you
define. This vignette walks through the workflow with stable innovations and
then builds a user defined distribution.

## The model

In the intercept form of Wuertz et al. (2006),

$$x_t = \mu + \sum_{i=1}^m a_i x_{t-i} + \sum_{j=1}^n b_j e_{t-j} + e_t, \qquad e_t = \sigma_t z_t,$$
$$\sigma_t^\delta = \omega + \sum_{i=1}^p \alpha_i (|e_{t-i}| - \gamma_i e_{t-i})^\delta + \sum_{j=1}^q \beta_j \sigma_{t-j}^\delta,$$

with $z_t$ i.i.d. from the innovation distribution, standardized to location
0 and scale 1. Stable innovations have infinite variance, so $\sigma_t$ is a
conditional *scale*, and $E|z|^\delta$ is finite only for $\delta < \alpha$.
In GARCH mode (`aparch = FALSE`) the power is fixed at 1 for stable
innovations and 2 for GEV and GAt.

## Stable AR(1)-APARCH(1,1)

A specification says which model to fit; a model adds parameter values.

```{r}
spec <- gs_spec(arma = c(1, 0), garch = c(1, 1), aparch = TRUE, dist = gs_stable())
spec
model <- gs_model(spec, mu = 0.0005, ar = 0.05, omega = 0.0002, alpha = 0.05,
                  gamma = 0.3, beta = 0.85, delta = 1.2,
                  dist_par = c(stable_alpha = 1.75, stable_beta = 0))
gs_stationarity(model)
```

The persistence $\alpha\, E(|z| - \gamma z)^\delta + \beta$ is below one, so
the model has a stationary scale level. Simulate and fit:

```{r}
x <- gs_sim(model, n = 1500, seed = 1)$y
fit <- gs_fit(x, spec)
fit
```

Three kinds of residuals are available. The raw residuals are
$e_t = x_t - \hat\mu - \sum \hat a_i x_{t-i} - \sum \hat b_j e_{t-j}$. Dividing them by the
fitted scale gives the standardized residuals $\hat z_t = e_t / \hat\sigma_t$, the
estimates of the innovations: if the model is right they are i.i.d. from the
fitted stable law, so they should show no remaining autocorrelation, in
themselves or in their absolute values (Ljung-Box tests).

Applying the fitted distribution function gives the probability integral
transform $\hat u_t = F(\hat z_t)$. Under a correct model $\hat u_t$ is i.i.d.
uniform on (0, 1), which gives a goodness of fit test of the innovation
distribution (Kolmogorov-Smirnov below), and these are the pseudo observations
that copula models take as input.

```{r}
z <- residuals(fit, type = "standardized")
Box.test(abs(z), lag = 20, type = "Ljung-Box")$p.value
u <- residuals(fit, type = "pit")
ks.test(u, "punif")$p.value
```

## Forecasting and value at risk

One step ahead the forecast is exact: location, scale and quantiles of a
shifted and scaled stable law. The value at risk at level $p$ is minus the $p$
quantile.

```{r}
predict(fit, n_ahead = 1, level = c(0.01, 0.05))
```

Beyond one step the scale follows the APARCH recursion and the quantiles
come from simulated paths.

```{r}
predict(fit, n_ahead = 5, level = 0.01, n_sim = 5000, seed = 2)
```

A rolling window backtest evaluates one step forecasts out of sample. The
window has a fixed length $w$. At each origin $t$ the model is fitted to the
last $w$ observations $x_{t-w+1}, \dots, x_t$ and the one step VaR for
$t + 1$ is computed. The window then moves forward by one: the *observed*
value $x_{t+1}$ (not the forecast) enters on the right and $x_{t-w+1}$ leaves
on the left, the model is refitted and the next VaR computed. Comparing each
VaR with the observation it was made for gives the violations used in the
Kupiec and Christoffersen tests.

Consecutive windows share $w - 1$ observations, so the estimates change little
from one origin to the next. Starting each fit from the previous estimates and
skipping the Hessian makes each refit several times faster:

```{r, eval = FALSE}
w <- 1000
origins <- w:(length(x) - 1)
var_1 <- numeric(length(origins)); prev <- NULL
for (k in seq_along(origins)) {
  t <- origins[k]
  fit_t <- gs_fit(x[(t - w + 1):t], spec, start = prev, hessian = FALSE)
  prev <- coef(fit_t)
  var_1[k] <- -predict(fit_t, level = 0.01)$q_0.01    # VaR for x[t + 1]
}
violations <- x[origins + 1] < -var_1
mean(violations)                                      # should be close to 0.01
```

## A user defined innovation distribution

Any distribution can be plugged in with `gs_dist()`. Every field that the
machinery needs is required, including when moments exist (`max_power`) and
the mean (`mean`, which must return `NA` explicitly if it does not exist).
The constructor checks the pieces against each other and stops with an error
naming the faulty field, so a mistake cannot silently distort a fit. Here the
normal distribution:

```{r}
norm <- gs_dist(
  name = "norm",
  log_density = function(z, par) dnorm(z, log = TRUE),
  random = function(n, par) rnorm(n),
  cdf = function(q, par) pnorm(q),
  quantile = function(p, par) qnorm(p),
  par_names = character(0), lower = numeric(0), upper = numeric(0), start = numeric(0),
  default_delta = 2,
  max_power = function(par) Inf,
  mean = function(par) 0)
```

A wrong declaration is caught immediately:

```{r, error = TRUE}
gs_dist(
  name = "norm",
  log_density = function(z, par) dnorm(z, log = TRUE),
  random = function(n, par) rnorm(n),
  cdf = function(q, par) pnorm(q),
  quantile = function(p, par) qnorm(p),
  par_names = character(0), lower = numeric(0), upper = numeric(0), start = numeric(0),
  default_delta = 2,
  max_power = function(par) Inf,
  mean = function(par) 0.5)
```

The new distribution gets the same fitting, stationarity and forecasting:

```{r}
spec_n <- gs_spec(garch = c(1, 1), dist = norm)
m_n <- gs_model(spec_n, mu = 0, omega = 0.1, alpha = 0.1, beta = 0.8)
fit_n <- gs_fit(gs_sim(m_n, n = 1000, seed = 3)$y, spec_n)
coef(fit_n)
predict(fit_n, n_ahead = 3, n_sim = 2000, seed = 4)
```

## Relationship to other packages

This package builds on existing work. The ARMA-APARCH filter and model
structure follow fGarch (Wuertz, Chalabi and Luksan, 2006); maximum
likelihood optimization uses Rsolnp (Galanos and Ye, 2025); and the stable
density uses libstable4u (Royuela-del-Val, Simmross-Wattenberg and
Alberola-Lopez, 2017). Its own contribution is the pluggable
innovation-distribution interface, the closed-form APARCH moments for the
stable, GEV and GAt families, and forecasting and Value-at-Risk for these
families.

## References

Galanos, A. and Ye, Y. (2025). Rsolnp: General Non-Linear Optimization. R
package version 2.0.1. <https://CRAN.R-project.org/package=Rsolnp>

Mittnik, S., Paolella, M. S. and Rachev, S. T. (2002). Stationarity of stable
power-GARCH processes. *Journal of Econometrics*, 106, 97-107.

Paolella, M. S. (1997). Tail estimation and conditional modeling of
heteroskedastic time series. PhD thesis, University of Kiel.

Royuela-del-Val, J., Simmross-Wattenberg, F. and Alberola-Lopez, C. (2017).
libstable: Fast, Parallel, and High-Precision Computation of alpha-Stable
Distributions in R, C/C++, and MATLAB. *Journal of Statistical Software*,
78(1), 1-25.

Wuertz, D., Chalabi, Y. and Luksan, L. (2006). Parameter estimation of ARMA
models with GARCH/APARCH errors: an R and SPlus software implementation.
*Journal of Statistical Software*.
